Possible Inconsistency in Maximum Likelihood Calculation of Ancestral States

### 2017-03-28 joint_inf.tex
Bringing commands into main text; easier to find.
updates

\input{commands}

@@

We will be considering phylogenetic likelihood on unrooted phylogenetic trees.
We assume the character alphabet $\alphabet=\{0,1\}$ and a uniform stationary distribution.
Let $m$ be the number of tips of the tree, and $p = m-2$ the number of internal nodes.
%EM Commented out below because it's not right-- fullAlignment is not the set of assignments of characters to tips of the tree, it's a single such.
%Let $\fullAlignment$ be the set of assignments of characters to tips of our tree, and $\fullAncestralStates$ the set of assignments of characters to the internal nodes of the tree.
Assume we observe $n$ samples of character data (i.e.\ an alignment with $n$ columns) $\fullAlignment=[\alignmentColumn_1,\ldots,\alignmentColumn_n]\in\alphabet^{\nSiteRows\times n}$.
There are also corresponding unknown ancestral states $\fullAncestralStates=[\ancestralStateColumn_1,\ldots,\ancestralStateColumn_n]\in\alphabet^{\nAncestralStateRows\times n}$.

@@

For a topology $\tau$ and branch lengths $t$ we form the fully-observed likelihood (making the usual independence assumption between sites):
\begin{equation}
\label{eq:full_likelihood}
L_n(\fullAlignment;\fullAncestralStates, \tau, t) = \prod_{i=1}^{n} \ P(\alignmentColumn_i, \ancestralStateColumn_i | \tau, t).
\end{equation}
% In particular, we are interested in
% $$
% (\hat{\boldsymbol\xi}, \hat{\tau}, \hat{t}) = \arg\max_{\boldsymbol\xi, \tau, t} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t),
% $$
% which we call the maximum likelihood values of the parameters $(\boldsymbol\xi, \tau, t)$.
% Since the number of elements in $\boldsymbol\xi$ grows with that of the observed data $\mathbf{y}$,
The typical approach to estimate the tree $\tau$ and branch lengths $t$ involves computing the marginal likelihood
\begin{equation}
\label{eq:marginal_likelihood}
\tilde{L}_n(\fullAlignment; \tau, t) = \sum_{\fullAncestralStates} \ L_n(\fullAlignment;\fullAncestralStates, \tau, t)
\end{equation}
and maximizing over the topology and branch lengths to obtain
$$
(\hat{\tau}, \hat{t}) = \arg\max_{\tau, t} \  \tilde{L}_n(\fullAlignment; \tau, t).
$$
% The values $\hat{\boldsymbol\xi}$ are then calculated conditional on these estimates.
% More likely in practice we fix a topology $\tau$ and use this marginalization approach to compute $(\hat{\boldsymbol\xi}, \hat{t})$.

@@

One tantalizing approach is to do away with the marginalization and directly estimate the maximum likelihood parameters from the full likelihood in \eqref{eq:full_likelihood}.
We can perform this by computing the profile likelihood
\begin{equation}
\label{eq:profile_likelihood}
L_n'(\fullAlignment;\tau, t) = \max_{\fullAncestralStates} \ L_n(\fullAlignment;\fullAncestralStates, \tau, t),
\end{equation}
then estimating the topology and branch lengths via
\begin{equation}
\label{eq:profile_likelihood_topology_bl}
(\hat{\tau}, \hat{t}) = \arg\max_{\tau, t} \ L_n'(\fullAlignment;\tau, t)
\end{equation}
while using $\hat{\fullAncestralStates}$ maximizing \eqref{eq:profile_likelihood} as an estimate for $\fullAncestralStates$.
Since $\alphabet$ is a discrete set, there exists a maximum \eqref{eq:profile_likelihood}, and so \eqref{eq:profile_likelihood_topology_bl} will recover the joint ML values of the unknown parameters.
In the general case, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities that depend on the unknown $(\tau,t)$.
For this reason, in practice the joint ML strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence cite{Neher2017}.

\subsection{Site pattern formulation}

Since we have a finite character alphabet, for a given column $i$ there are a finite number of possibilities of assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$.
These are commonly called ``site patterns'' in the literature.
%EM in general aren't a different number of internal site patterns than tip site patterns?
%das: Yes, I moved the paragraph at the end to here so it's hopefully less confusing. Maybe writing out the example would help? I also changed from "pattern" to "category" which might make it a little more clear.
%EM Seems good this is an important concept so it's worth defining it more properly.
We enumerate the unique site patterns $\mathbf{s}=[s_1,\ldots,s_q]\in\mathcal{S}$; these are defined by the number of leaves and the character alphabet.
Corresponding to each of these site patterns will be an ancestral state category that is most likely given a particular likelihood.
Since $\alphabet$ is a finite set, the value of each factor $P(\alignmentColumn_i, \ancestralStateColumn_i | \tau, t)$ in \eqref{eq:full_likelihood} is maximized at one or more values of $\ancestralStateColumn_i$, though independent of $i$ given $j$.
This independence allows us to group factors by those where $\ancestralStateColumn_i$ belongs to an ancestral state category $\ancestralStateCategory_j$ so that, if $\alignmentColumn_i=\sitePattern_j$ and $\ancestralStateColumn_i\in \ancestralStateCategory_j$, $P(\sitePattern_j, \ancestralStateCategory_j | \tau, t)$ is maximized.

For ancestral state categories $\fullAncestralStateCategories=[\ancestralStateCategory_1,\ldots,\ancestralStateCategory_q]\in\ancestralStateCategorySpace$ we reformulate the likelihood equation
\begin{align}
L_n'(\fullAlignment;\tau, t) &= \max_{\fullAncestralStates} \ L_n(\fullAlignment;\fullAncestralStates, \tau, t) \\
                             &= \max_{\fullAncestralStates} \ \prod_{i=1}^{n} \ P(\alignmentColumn_i, \ancestralStateColumn_i | \tau, t) \\
%EM Can you think a little about z_j?
%EM Even though it's a category now, it's still a function of tau and t, right?
                             &= \max_{\fullAncestralStateCategories} \ \prod_{j=1}^{q} \ P(\sitePattern_j, \ancestralStateCategory_j | \tau, t)^{n_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
where $n_j(\fullAlignment)$ is the number of observations in the sample where site pattern $\sitePattern_j$ was seen in $\fullAlignment$.

\textbf{Example:} This might be a good time for a toy example showing this for a very simple topology.

@@

We are now interested in the properties of the objective function being maximized in \eqref{eq:site_pattern_likelihood} as data becomes plentiful.
Let
$$
L_n''(\fullAlignment;\fullAncestralStateCategories,\tau,t) = \prod_{j=1}^q \ P(\sitePattern_j, \ancestralStateCategory_j | \tau, t)^{n_j(\fullAlignment)}
$$
so that
$$
\max_{\fullAncestralStates} \ L_n(\fullAlignment;\fullAncestralStates, \tau, t) =
    \max_{\fullAncestralStateCategories} \ L_n''(\fullAlignment;\fullAncestralStateCategories,\tau,t).
$$
Assume our $n$ observations were generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{align}
    \frac{1}{n} \log L_n''(\fullAlignment;\fullAncestralStateCategories,\tau,t)
        &= \sum_{j} \frac{n_j(\fullAlignment)}{n}\cdot  \log P(\sitePattern_j, \ancestralStateCategory_j | \tau, t) \\
        &= \sum_{j} \frac{n_j(\fullAlignment)}{n}\cdot [\log P(\sitePattern_j | \tau, t) + \log P(\ancestralStateCategory_j | \sitePattern_j, \tau, t)].
\end{align}
In the limit of infinite data we have
$$
\frac{1}{n} \log L_n''(\fullAlignment;,\tau,t) \rightarrow \sum_{j} P(\sitePattern_j | \tau^*, t^*) \cdot [\log P(\sitePattern_j | \tau, t) + \log P(\ancestralStateCategory_j | \sitePattern_j, \tau, t)]
$$
and that, since $\ancestralStateCategory_j$ appears only in the last term, the log of \eqref{eq:site_pattern_likelihood} in the same limit is proportional to
\begin{align}
\frac{1}{n} \log L_n'(\fullAlignment;\fullAncestralStateCategories,\tau, t)
    &= \max_{\fullAncestralStateCategories} \ \frac{1}{n} \log L_n''(\fullAlignment;\fullAncestralStateCategories,\tau,t) \nonumber \\
    &\rightarrow \max_{\fullAncestralStateCategories} \ \sum_{j} P(\sitePattern_j | \tau^*, t^*) \cdot [\log P(\sitePattern_j | \tau, t) + \log P(\ancestralStateCategory_j | \sitePattern_j, \tau, t)] \nonumber \\
    &= \max_{\fullAncestralStateCategories} \ \sum_{j} P(\sitePattern_j | \tau^*, t^*) \cdot [\log P(\sitePattern_j | \tau, t) + \log P(\ancestralStateCategory_j | \sitePattern_j, \tau, t)] \nonumber \\
    &= \max_{\fullAncestralStateCategories} \ \left\{\sum_{j} P(\sitePattern_j | \tau^*, t^*) \cdot \left[\log P(\sitePattern_j | \tau, t) + \log P(\ancestralStateCategory_j | \sitePattern_j, \tau, t)\right]\right\}. \label{eq:site_pattern_profile_likelihood_mean}
\end{align}
Define the the information quantity
$$
H_{\tau^*,t^*}(\tau,t) = \sum_{j} P(\sitePattern_j | \tau^*, t^*) \log P(\sitePattern_j | \tau, t),
$$
and the partial log likelihood
$$
\ell^p_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) = \sum_{j} P(\sitePattern_j | \tau^*, t^*) \log P(\ancestralStateCategory_j | \sitePattern_j, \tau, t)
$$
so that \eqref{eq:site_pattern_profile_likelihood_mean} is written as
\begin{equation}
    \label{eq:log_likelihood_simplified}
\ell_{\tau^*,t^*}(\tau,t) = \max_{\fullAncestralStateCategories} \ \left\{H_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t)\right\}.
\end{equation}
Recall $(\tau, t)$ are estimands and $(\tau^*, t^*)$ are fixed, generating parameters.
For data generated from $(\tau^*, t^*)$, if
\begin{equation}
\label{eq:inconsistency_inequality}
\ell_{\tau^*,t^*}(\tau',\hat{t}_1) > \ell_{\tau^*,t^*}(\tau^*,\hat{t}_2)
\end{equation}
for $\tau'\neq\tau^*$ with $\hat{t}_1$ and $\hat{t}_2$ estimated by maximizing \eqref{eq:profile_likelihood}, we have an inconsistency.

@@

Our approach will be to seek bounds for the terms inside the maximization in \eqref{eq:log_likelihood_simplified} as
\begin{equation}
\label{eq:partial-likelihood}
\ell_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) = H_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t).
\end{equation}
We obtain an upper bound for the joint maximum of \eqref{eq:partial-likelihood}
$$
\max_{\{t,\fullAncestralStateCategories\}} \ \ell_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) \le
    H_{\tau^*,t^*}(\tau^*,t^*)
    + \ell^p_{\tau^*,t^*}(\hat{\fullAncestralStateCategories},\tau,\hat{t})
$$
using Gibbs's inequality
$$
H_{\tau^*,t^*}(\tau,t) \le H_{\tau^*,t^*}(\tau^*,t^*)
$$
and
$$
\{\hat{\fullAncestralStateCategories},\hat{t}\} = \arg\max_{\{\fullAncestralStateCategories,t\}} \ \ell^p_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t).
$$
Similarly, we have a lower bound as
$$
\max_{\{t,\fullAncestralStateCategories\}} \ \ell_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) \ge
    H_{\tau^*,t^*}(\tau,\hat{t})
    + \ell^p_{\tau^*,t^*}(\hat{\fullAncestralStateCategories},\tau,\hat{t}).
$$

@@

\begin{table}
\centering
\begin{tabular}{|l|l|l|}
\multicolumn{3}{c}{Classical}\\
    \hline
$\sitePattern_j$   &$P(\sitePattern_j|\tau_1,t^*)$&$P(\sitePattern_j|\tau_2,t^*)$\\
    \hline
0000&$1+x^2+y^2+4xy^2+x^2y^2$&$1+2xy+2xy^2+x^2y+y^3+x^2y^2$\\
0001&$1+x^2-y^2-x^2y^2$&$1+x^2y-y^3-x^2y^2$\\
0010&$1-x^2+y^2-x^2y^2$&$1-x^2y+y^3-x^2y^2$\\
0100&$1+x^2-y^2-x^2y^2$&$1+x^2y-y^3-x^2y^2$\\
1000&$1-x^2+y^2-x^2y^2$&$1-x^2y+y^3-x^2y^2$\\
0011&$1-x^2-y^2+x^2y^2$&$1+2xy-2xy^2-x^2y-y^3+x^2y^2$\\
0101&$1+x^2+y^2-4xy^2+x^2y^2$&$1-2xy-2xy^2+x^2y+y^3+x^2y^2$\\
1001&$1-x^2-y^2+x^2y^2$&$1-2xy+2xy^2-x^2y-y^3+x^2y^2$\\
    \hline
\multicolumn{3}{c}{Modified}\\
    \hline
$\sitePattern_j$   &$P(\sitePattern_j|\tau_1,t)$&$P(\sitePattern_j|\tau_2,t)$\\
    \hline
0000&$1+x^2+y^2+4xyw+x^2y^2$&$1+2xy+2xyw+x^2w+y^2w+x^2y^2$\\
0001&$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
0010&$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
0100&$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
1000&$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
0011&$1-x^2-y^2+x^2y^2$&$1+2xy-2xyw-x^2w-y^2w+x^2y^2$\\
0101&$1+x^2+y^2-4xyw+x^2y^2$&$1-2xy-2xyw+x^2w+y^2w+x^2y^2$\\
1001&$1-x^2-y^2+x^2y^2$&$1-2xy+2xyw-x^2w-y^2w+x^2y^2$\\
    \hline
\end{tabular}
\caption{Site pattern probabilities.
All values are multiplied by $1/8$.}
\label{tab:sitepatprob}
\end{table}

>>>>

% Commands
\newcommand{\alphabet}{\mathcal{A}}
\newcommand{\fullAlignment}{\mathbf{Y}}
\newcommand{\alignmentColumn}{\mathbf{y}}
\newcommand{\sitePattern}{s}
\newcommand{\siteSplit}{S'}
\newcommand{\sitePatternSet}{S}
\newcommand{\fullSitePatterns}{\mathbf{s}}
\newcommand{\sitePatternSpace}{\mathcal{S}}
\newcommand{\fullAncestralStates}{\mathbf{H}}
\newcommand{\ancestralStateSpace}{\mathcal{H}}
\newcommand{\ancestralStateColumn}{\boldsymbol\eta}
\newcommand{\ancestralStateCategory}{Z}
\newcommand{\ancestralSplit}{H'}
\newcommand{\ancestralStateSet}{H}
\newcommand{\fullAncestralStateCategories}{\mathbb{Z}}
\newcommand{\ancestralStateCategorySpace}{\mathcal{Z}}

\newcommand{\nCols}{n}
\newcommand{\nSiteRows}{m}
\newcommand{\nAncestralStateRows}{p}
\newcommand{\nSiteSplits}{q}
\newcommand{\nAncestralSplits}{r}
\newcommand{\nMostLikelyAncestralSplits}{\tilde{r}}

@@

We will be considering phylogenetic likelihood on unrooted phylogenetic trees.
We assume the character alphabet $\alphabet=\{0,1\}$ and a uniform stationary distribution.
Let $\nSiteRows$ be the number of tips of the tree, and $\nAncestralStateRows = \nSiteRows-2$ the number of internal nodes.
%EM Commented out below because it's not right-- fullAlignment is not the set of assignments of characters to tips of the tree, it's a single such.
%Let $\fullAlignment$ be the set of assignments of characters to tips of our tree, and $\fullAncestralStates$ the set of assignments of characters to the internal nodes of the tree.
Assume we observe $\nCols$ samples of character data (i.e.\ an alignment with $\nCols$ columns) $\fullAlignment=[\alignmentColumn_1,\ldots,\alignmentColumn_\nCols]\in\alphabet^{\nSiteRows\times\nCols}$.
There are also corresponding unknown ancestral states $\fullAncestralStates=[\ancestralStateColumn_1,\ldots,\ancestralStateColumn_\nCols]\in\alphabet^{\nAncestralStateRows\times\nCols}$.

@@

For a topology $\tau$ and branch lengths $t$ we form the fully-observed likelihood (making the usual independence assumption between sites):
\begin{equation}
\label{eq:full_likelihood}
L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t) = \prod_{i=1}^{\nCols} \ P(\alignmentColumn_i, \ancestralStateColumn_i | \tau, t).
\end{equation}
% In particular, we are interested in
% $$
% (\hat{\boldsymbol\xi}, \hat{\tau}, \hat{t}) = \arg\max_{\boldsymbol\xi, \tau, t} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t),
% $$
% which we call the maximum likelihood values of the parameters $(\boldsymbol\xi, \tau, t)$.
% Since the number of elements in $\boldsymbol\xi$ grows with that of the observed data $\mathbf{y}$,
The typical approach to estimate the tree $\tau$ and branch lengths $t$ involves computing the marginal likelihood
\begin{equation}
\label{eq:marginal_likelihood}
\tilde{L}_\nCols(\fullAlignment; \tau, t) = \sum_{\fullAncestralStates} \ L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t)
\end{equation}
and maximizing over the topology and branch lengths to obtain
$$
(\hat{\tau}, \hat{t}) = \arg\max_{\tau, t} \  \tilde{L}_\nCols(\fullAlignment; \tau, t).
$$
% The values $\hat{\boldsymbol\xi}$ are then calculated conditional on these estimates.
% More likely in practice we fix a topology $\tau$ and use this marginalization approach to compute $(\hat{\boldsymbol\xi}, \hat{t})$.

@@

One tantalizing approach is to do away with the marginalization and directly estimate the maximum likelihood parameters from the full likelihood in \eqref{eq:full_likelihood}.
We can perform this by computing the profile likelihood
\begin{equation}
\label{eq:profile_likelihood}
L_\nCols'(\fullAlignment;\tau, t) = \max_{\fullAncestralStates} \ L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t),
\end{equation}
then estimating the topology and branch lengths via
\begin{equation}
\label{eq:profile_likelihood_topology_bl}
(\hat{\tau}, \hat{t}) = \arg\max_{\tau, t} \ L_\nCols'(\fullAlignment;\tau, t)
\end{equation}
while using $\hat{\fullAncestralStates}$ maximizing \eqref{eq:profile_likelihood} as an estimate for $\fullAncestralStates$.
Since $\alphabet$ is a discrete set, there exists a maximum \eqref{eq:profile_likelihood}, and so \eqref{eq:profile_likelihood_topology_bl} will recover the joint ML values of the unknown parameters.
In the general case, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities that depend on the unknown $(\tau,t)$.
For this reason, in practice the joint ML strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \cite{Neher2017}.

\subsection{Site split formulation}

Since we have a finite character alphabet, for a given column $i$ there are a finite number of possibilities of assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$.
Following Section 8.6 of \cite{Semple2003}, $\tau$ induces an ordered set of tip labels $\{1,\ldots,\nSiteRows\}$ across sites.
For a binary character model, a realization of $\alignmentColumn_i$ can be described as a subset of complementary characters $\siteSplit\subseteq\sitePatternSet\triangleq\{1,\ldots,\nSiteRows-1\}$.
The topology $\tau$ induces an ordered set $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ with a similar subset of complementary characters $\ancestralSplit\subseteq\ancestralStateSet\triangleq\{1,\ldots,\nAncestralStateRows\}$ describing a realization of $\ancestralStateColumn_i$.
Define the total number of unique site splits as $\nSiteSplits=|\mathcal{P}(\sitePatternSet)|$ and unique ancestral state splits as $\nAncestralSplits=|\mathcal{P}(\ancestralStateSet)|$ where $\mathcal{P}$ denotes the powerset.

%EM in general aren't a different number of internal site patterns than tip site patterns?
%das: Yes, I moved the paragraph at the end to here so it's hopefully less confusing. Maybe writing out the example would help? I also changed from "pattern" to "category" which might make it a little more clear.
%EM Seems good this is an important concept so it's worth defining it more properly.

Consider the probability $P(\alignmentColumn_i, \ancestralStateColumn_i | \tau, t)$ from \eqref{eq:full_likelihood} of observing a particular column $i$.
Given $(\tau, t)$ and maximizing over $\ancestralStateColumn_i$, if $\alignmentColumn_i=\siteSplit_j$ then there exists a partition $\ancestralStateCategory_j$ of $\mathcal{P}(\ancestralStateSet)$ of most likely ancestral states where
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralStateSet)} \ P(\alignmentColumn_i=\siteSplit_j, \ancestralStateColumn_i=\ancestralSplit_k | \tau, t) &= P(\alignmentColumn_i=\siteSplit_j, \ancestralStateColumn_i\in\ancestralStateCategory_j | \tau, t) \\
     &= P(\alignmentColumn=\siteSplit_j, \ancestralStateColumn\in\ancestralStateCategory_j | \tau, t).
\end{align*}
This equivalence is helpful in writing the full likelihood of the joint inference problem in terms of site splits; for most likely ancestral state partitions $\fullAncestralStateCategories=\{\ancestralStateCategory_1,\ldots,\ancestralStateCategory_\nSiteSplits\}$ we reformulate the likelihood equation
\begin{align}
L_\nCols'(\fullAlignment;\tau, t) &= \max_{\fullAncestralStates} \ L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t) \\
                             &= \max_{\fullAncestralStates} \ \prod_{i=1}^{\nCols} \ P(\alignmentColumn_i, \ancestralStateColumn_i | \tau, t) \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralStateSet)} \ P(\alignmentColumn_i=\siteSplit_j, \ancestralStateColumn_i=\ancestralSplit_k | \tau, t) \\
%EM Can you think a little about z_j?
%EM Even though it's a category now, it's still a function of tau and t, right?
                             &= \prod_{i=1}^{\nCols} \ P(\alignmentColumn=\siteSplit_j, \ancestralStateColumn\in\ancestralStateCategory_j | \tau, t) \\
                             &= \prod_{j=1}^{\nSiteSplits} \ P(\alignmentColumn=\siteSplit_j, \ancestralStateColumn\in\ancestralStateCategory_j | \tau, t)^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ where the site split $\siteSplit_j$ appears.

@@

We are interested in the properties of the objective function being maximized in \eqref{eq:site_pattern_likelihood} as data becomes plentiful.
Define \eqref{eq:site_pattern_likelihood} as
$$
L_\nCols''(\fullAlignment;\fullAncestralStateCategories,\tau,t) = \prod_{j=1}^{\nSiteSplits} \ P(\alignmentColumn=\siteSplit_j, \ancestralStateColumn\in\ancestralStateCategory_j | \tau, t)^{\nCols_j(\fullAlignment)} .
$$
Assume our $\nCols$ observations were generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{align}
    \frac{1}{\nCols} \log L_\nCols''(\fullAlignment;\fullAncestralStateCategories,\tau,t)
        &= \sum_{j}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\alignmentColumn=\siteSplit_j, \ancestralStateColumn\in\ancestralStateCategory_j | \tau, t) \\
        &= \sum_{j}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\alignmentColumn=\siteSplit_j | \tau, t) + 
            \log P(\ancestralStateColumn\in\ancestralStateCategory_j | \alignmentColumn=\siteSplit_j , \tau, t)]
\end{align}
so that, in the limit of infinite data,
\begin{equation}
\frac{1}{\nCols} \log L_\nCols''(\fullAlignment;,\tau,t) \rightarrow \sum_{j}^\nSiteSplits P(\alignmentColumn=\siteSplit_j | \tau^*, t^*) \cdot [\log P(\alignmentColumn=\siteSplit_j | \tau, t) + \log P(\ancestralStateColumn\in\ancestralStateCategory_j | \alignmentColumn=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{equation}
Define the the information quantity
$$
h_{\tau^*,t^*}(\tau,t) = \sum_{j}^\nSiteSplits P(\alignmentColumn=\siteSplit_j | \tau^*, t^*) \log P(\siteSplit_j | \tau, t),
$$
and the partial log likelihood
$$
\ell^p_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) = \sum_{j}^\nSiteSplits P(\alignmentColumn=\siteSplit_j | \tau^*, t^*) \log P(\ancestralStateColumn\in\ancestralStateCategory_j | \siteSplit_j, \tau, t)
$$
so that \eqref{eq:site_pattern_profile_likelihood_mean} is written as
\begin{equation}
    \label{eq:log_likelihood_simplified}
\ell_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) = h_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t).
\end{equation}
Recall $(\tau, t)$ are estimands and $(\tau^*, t^*)$ are fixed, generating parameters.
For data generated from $(\tau^*, t^*)$, if
\begin{equation}
\label{eq:inconsistency_inequality}
\ell_{\tau^*,t^*}(\hat{\fullAncestralStateCategories}_1,\tau',\hat{t}_1) > \ell_{\tau^*,t^*}(\hat{\fullAncestralStateCategories}_2,\tau^*,\hat{t}_2)
\end{equation}
for $\tau'\neq\tau^*$ with $\{\hat{t}_1,\hat{\fullAncestralStateCategories}_1\}$ and $\{\hat{t}_2,\hat{\fullAncestralStateCategories}_2\}$ estimated by maximizing \eqref{eq:profile_likelihood}, we have an inconsistency.

@@

Our approach will be to seek bounds for the terms inside the maximization in \eqref{eq:log_likelihood_simplified} as
\begin{equation}
\label{eq:partial-likelihood}
\ell_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) = h_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t).
\end{equation}
We obtain an upper bound for the joint maximum of \eqref{eq:partial-likelihood}
$$
\max_{\{t,\fullAncestralStateCategories\}} \ \ell_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) \le
    h_{\tau^*,t^*}(\tau^*,t^*)
    + \ell^p_{\tau^*,t^*}(\hat{\fullAncestralStateCategories},\tau,\hat{t})
$$
using Gibbs's inequality
$$
h_{\tau^*,t^*}(\tau,t) \le h_{\tau^*,t^*}(\tau^*,t^*)
$$
and
$$
\{\hat{\fullAncestralStateCategories},\hat{t}\} = \arg\max_{\{\fullAncestralStateCategories,t\}} \ \ell^p_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t).
$$
Similarly, we have a lower bound as
$$
\max_{\{t,\fullAncestralStateCategories\}} \ \ell_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) \ge
    h_{\tau^*,t^*}(\tau,\hat{t})
    + \ell^p_{\tau^*,t^*}(\hat{\fullAncestralStateCategories},\tau,\hat{t}).
$$

@@

\begin{table}
\centering
\begin{tabular}{|l|l|l|}
\multicolumn{3}{c}{Classical}\\
    \hline
$\siteSplit_j$   &$P(\alignmentColumn=\siteSplit_j|\tau_1,t^*)$&$P(\alignmentColumn=\siteSplit_j|\tau_2,t^*)$\\
    \hline
    $\emptyset$&$1+x^2+y^2+4xy^2+x^2y^2$&$1+2xy+2xy^2+x^2y+y^3+x^2y^2$\\
    $\{1\}$    &$1-x^2+y^2-x^2y^2$&$1-x^2y+y^3-x^2y^2$\\
    $\{2\}$    &$1+x^2-y^2-x^2y^2$&$1+x^2y-y^3-x^2y^2$\\
    $\{3\}$    &$1-x^2+y^2-x^2y^2$&$1-x^2y+y^3-x^2y^2$\\
    $\{1,2\}$  &$1-x^2-y^2+x^2y^2$&$1+2xy-2xy^2-x^2y-y^3+x^2y^2$\\
    $\{1,3\}$  &$1+x^2+y^2-4xy^2+x^2y^2$&$1-2xy-2xy^2+x^2y+y^3+x^2y^2$\\
    $\{2,3\}$  &$1-x^2-y^2+x^2y^2$&$1-2xy+2xy^2-x^2y-y^3+x^2y^2$\\
    $\{1,2,3\}$&$1+x^2-y^2-x^2y^2$&$1+x^2y-y^3-x^2y^2$\\
    \hline
\multicolumn{3}{c}{Modified}\\
    \hline
$\siteSplit_j$   &$P(\alignmentColumn=\siteSplit_j|\tau_1,t)$&$P(\alignmentColumn=\siteSplit_j|\tau_2,t)$\\
    \hline
    $\emptyset$&$1+x^2+y^2+4xyw+x^2y^2$&$1+2xy+2xyw+x^2w+y^2w+x^2y^2$\\
    $\{1\}$    &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
    $\{2\}$    &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
    $\{3\}$    &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
    $\{1,2\}$  &$1-x^2-y^2+x^2y^2$&$1+2xy-2xyw-x^2w-y^2w+x^2y^2$\\
    $\{1,3\}$  &$1+x^2+y^2-4xyw+x^2y^2$&$1-2xy-2xyw+x^2w+y^2w+x^2y^2$\\
    $\{2,3\}$  &$1-x^2-y^2+x^2y^2$&$1-2xy+2xyw-x^2w-y^2w+x^2y^2$\\
    $\{1,2,3\}$&$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
    \hline
\end{tabular}
\caption{Site pattern probabilities.
All values are multiplied by $1/8$.}
\label{tab:sitepatprob}
\end{table}
